---
output:
  github_document:
    html_preview: false
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

# Shiny standalone (refreeze)
#### Notes and resources for how to create (*freeze*) a standalone shiny application

## Introduction

The [RStudio Shiny](https://shiny.rstudio.com/) web development framework provides a great set of tools to quickly create web applications for scientific and data management use. It is especially good for building user-friendly one-off applications to accomplish specific tasks such as exploring, analyzing or validating data...or producing estimates and reports. 

Often the main roadblock when creating shiny applications is not development, but deployment. Shiny webapps can always be run locally from the command line or [RStudio](https://www.rstudio.com/), but intended users are typically non-programmers, and may be uncomfortable running scripts. Deploying to low-cost services such as [shinyapps.io](https://www.shinyapps.io/) may not be an option if local databases need to be accessed. Within organizations, deploying small, non-standard projects to a webserver may be difficult or impossible given the inevitable hurdles of IT bureaucracy, security issues, or procurement. When the intended user base is relatively small, creating a standalone program may be the simplest and fastest way to get needed tools into the hands of users...especially when those programs do not require the end-user to have administrative rights in order to install on their local machines. 

Unlike scripts that can be subject to errors as the underlying R engine and dependent package universe changes, a standalone application is frozen in time. It will always work as when first constructed. Executable installers cost essentially nothing to copy and distribute. Scaling the user base is a non-issue. 

## Roadmap

The following steps document *one* simple and opinionated way to freeze shiny applications for local installs on Windows machines as standalone applications. I have successfully used this method since 2014 and am documenting the steps mostly for myself to avoid forgetting the procedure and repeating common mistakes. It is based on this [blog post](http://blog.analytixware.com/2014/03/packaging-your-shiny-app-as-windows.html) from 2014 when the shiny framework was still in early development. Next steps are to test and document procedures on how to package standalone shiny apps using [Electron](https://electronjs.org/) on the front-end. See: [r-shiny-electron](https://github.com/dirkschumacher/r-shiny-electron), [electricShine](https://github.com/chasemc/electricShine), [deploy-shiny-electron](https://www.travishinkelman.com/post/deploy-shiny-electron/), [electron-quick-start](https://github.com/ColumbusCollaboratory/electron-quick-start).  

### Project setup

* Create a new shiny application in an RStudio project. Test thorougly to make sure everything works as intended. Include [validate()](https://shiny.rstudio.com/reference/shiny/1.0.4/validate.html) and [need()](https://shiny.rstudio.com/articles/validation.html) functions liberally to avoid the possibility your application may crash after it has been packaged as a standalone program. Crashes typically require that the offending R process be killed using Windows Task Manager before the program can be restarted...not ideal for end-users. In practice, this is exceedingly rare as long as basic best practices are followed.

* I normally copy finalized shiny code intended to run standalone over to a separate applications directory. Create a new folder in your applications working directory. In my case this would be `C:\Documents\RStudio\Apps`. For the `flight_proof` application used as an example below, the path would be: `C:\Documents\RStudio\Apps\flight_proof`  

* Download the latest *suitable* version of [R-Portable](https://sourceforge.net/projects/rportable/files/R-Portable/) to the new folder. Then double-click the (currently) `R-Portable_3.6.3.paf.exe` to run. This will create the `R-Portable` folder in your new application folder. No admin privileges are needed to install `R-Portable`. Although at this writing there is an existing `R-Portable_4.0.0_paf.exe`, that version appears to have issues with the Rcpp package. Consequently, until an newer version of R-Portable is released, use the one listed above.  

* If a 32-bit version of R is needed to run your application, for example when a 32-bit MS Access database needs to be accessible, do not click `Yes` to the `Run R Portable` checkbox during the installation process...as the default assumes you want 64-bit R. 

* Install needed R packages, and package dependencies. Normally you would install any required packages by running the `Rgui.exe` located at `R-Portable\App\R-Portable\bin\x64`. For the less common cases where 32-bit R is needed I normally navigate to the `R-Portable\App\R-Portable\bin\i386` folder and then click on `Rgui.exe` to open the Rgui. I then install any needed packages using 32-bit R. It may not make any difference either way.  

* Using the Rgui (whether 64-bit or 32-bit) open the `library_install.R` script located in the [refreeze](https://github.com/arestrom/refreeze) repository. You will also need the `find_package_dependencies.R` script to do a deep recursive search of all required packages and dependencies. Copies of these scripts are also normally kept in the `RStudio\Apps` folder. Run the dependencies script to identify dependencies, then edit the package installation script as needed to add required libraries.

* Copy your `ui.R`, `server.R`, and `global.R` shiny scripts, or any other scripts needed for your application, into a folder named `shiny` in the top-level directory of your application. For example: `C:\Documents\RStudio\Apps\flight_proof\shiny`. There should also be a `www` subfolder containing any images, rmarkdown .rmd files, etc., directly under the `shiny` folder. 

* Make sure and edit the `server.R` file to uncomment the `stopApp()` function at the very bottom of the `server.R` script. This is needed to make sure the `R.exe` is properly closed when the application exits. It is normally commented out during development. The function looks like this:

```{r, eval = FALSE}
# close the R session when the browser closes
  session$onSessionEnded(function() {
    stopApp()
    q("no")
  })
```

* The [refreeze](https://github.com/arestrom/refreeze) repository contains example materials needed for the next steps. Copy the `run.vbs`, `runShinyApp.R`, and `FlightProof.ico` from the repository into the top level of your app folder. You should also copy the `flight_proof.iss` script that will be needed to create the setup executable. I normally keep a copy of all application `.iss` scripts in the applications parent directory, in my case `C:\Documents\RStudio\Apps`.  

* Rename the `FlightProof.ico` as needed, or find another icon to use. There should be no need to edit the other files. You will only need to edit `run.vbs` if 32-bit R is needed. In that case the text of the `run.vbs` script will be `R-Portable\App\R-Portable\bin\i386\R.exe` instead of `R-Portable\App\R-Portable\bin\x64\R.exe`. 

### Create Windows installer

* Download and install [Inno Setup](http://www.jrsoftware.org/isinfo.php). Then edit a copy of the `flight_proof.iss` Inno Setup script, and rename as needed to your new application name, for example, `app_name.iss`. 

* Double-click on the `app_name.iss` script to open the `Inno Setup` program. In the `AppID` field, highlight all but the first curly brace, and under `Tools` click on the `Generate GUID` option. This will overwrite the section with a new GUID. Every time you create a new version, a new GUID should be generated. Edit the other fields as needed to specify things such as the application name and version number, the source directory for the application files, and the destination directory for the setup executable. Make sure that the `PrivilegesRequired` field is set to `none`. This will ensure that end-users can run the setup executable without needing administrative privileges. In practice I have found that the setup dialogue may still ask for an administrative login in order to run the installer, but your normal login account name and password will work. 

* To create the application setup installer, click on `Build – Compile`. This will bundle up all materials needed for you application and create an executable installer in your application directory. Your application, along with the current version of R and all needed packages, are now frozen in time. 

### Connecting to databases

* When needing to connect to databases, providing connection credentials requires special handling. A secure and relatively foolproof method is to use the `odbc` package, and then set up each user with a DSN. This can be done by typing `odbc` into the Windows search bar, then clicking on the `ODBC Data Sources` application to set up a `User DSN`. Typically, this will be a 64-bit DSN. Connecting via ODBC to a `PostgreSQL` database means that you will also need to have the [PostgreSQL ODBC driver](https://odbc.postgresql.org/) installed on your computer. 

* If you are trying to connect to a `PostgreSQL` database, my current favorite method is to use the `RPostgres` package. This may change in the future, but for now it provides better handling of uuids and odd text encodings. It can also be much faster when writing large files. `RPostgres` includes it's own driver, so no separate driver needs to be installed. 

When connecting via `RPostgres` in `RStudio` one way to provide credentials is enter the values in the `.Renviron` file and use the `Sys.getenv()` function to pull them out. But this will not work with the current standalone setup outlined above. It may be possible, but so far I have not figured out how. It's easy to change the flags in the `run.vbs` file from `--vanilla` (does not read `.Renviron`) to one where the `.Renviron` is read, but the path to `HOME` is slightly different when running from `R.exe` in batch mode. If anyone knows how to get `R.exe` to read the correct path and return manually entered credentials (other stuff like "USERNAME" or "HOME" is properly read), I would love to know.

For now I am focusing on using the [shinyauthr](https://github.com/PaulC91/shinyauthr) package to manually enter passwords via the interface. This avoids having to hard-code credentials anywhere, and is probably the next best option to using DSNs.  

### In case of failure  

* If after installing the application, it crashes, or fails to load, navigate to the `out.txt` folder in the App directory that the installer created, for example: `C:\Documents\Intertidal\Apps\FlightProof\out.txt`. The `out.txt` file will log the first fatal error encountered. In the case of missing dependencies it will identify any packages that you may still be missing. Each missing dependency will trigger a crash, so in the past, before using the `find_package_dependencies.R` script, multiple iterations were typically needed to identify all missing package dependencies. Hopefully, this will now be a rare event.  

* In case the application crashes or fails to load, you will also most likely need to kill any R processes still active using Windows Task Manager. You will then need to fix the source of the crash, usually a missing package dependency, and recompile the application. 


