---
output:
  github_document:
    html_preview: false
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

# Shiny standalone (refreeze)
#### Notes and resources for how to create (*freeze*) a standalone shiny application

## Introduction

The [RStudio Shiny](https://shiny.rstudio.com/) web development framework provides a great set of tools to quickly create web applications for scientific and data management use. It is especially good for building user-friendly one-off applications to accomplish specific tasks such as exploring, analyzing or validating data...or producing estimates and reports. 

Often the main roadblock when creating shiny applications is not development, but deployment. Shiny webapps can always be run locally from the command line or [RStudio](https://www.rstudio.com/), but intended users are typically non-programmers, and may be uncomfortable running scripts. Deploying to low-cost services such as [shinyapps.io](https://www.shinyapps.io/) may not be an option if local databases need to be accessed. Within organizations, deploying small, non-standard, projects to a webserver may be difficult or impossible given the inevitable hurdles of IT bureaucracy, security issues, or procurement. When the intended user base is relatively small, creating a standalone program may be the simplest and fastest way to get needed tools into the hands of users. Especially when those programs do not require the end-user to have administrative rights in order to install on their local machines. 

Unlike scripts that can be subject to errors as the underlying R engine and dependent package universe changes, a standalone application is frozen in time. It will always work as when first constructed. Executable installers cost essentially nothing to copy and distribute. Scaling the user base is a non-issue. 

## Roadmap

The following steps document *one* simple and opinionated way to freeze shiny applications for local installs on Windows machines as standalone applications. I have successfully used this method since 2014 and am documenting the steps mostly for myself to avoid forgetting the procedure and repeating common mistakes. It is based on this [blog post](http://blog.analytixware.com/2014/03/packaging-your-shiny-app-as-windows.html) from 2014 when the shiny framework was still in early development. Next steps are to test and document procedures on how to package standalone shiny apps using [Electron](https://electronjs.org/) on the front-end. See: [r-shiny-electron](https://github.com/dirkschumacher/r-shiny-electron), [Rinno](https://ficonsulting.github.io/RInno/), [deploy-shiny-electron](https://www.travishinkelman.com/post/deploy-shiny-electron/), [electron-quick-start](https://github.com/ColumbusCollaboratory/electron-quick-start).  

### Project setup

* Create a new shiny application in an RStudio project. Test thorougly to make sure everything works as intended. Include [validate()](https://shiny.rstudio.com/reference/shiny/1.0.4/validate.html) and [need()](https://shiny.rstudio.com/articles/validation.html) functions liberally to avoid the possibility your application may crash after it has been packaged as a standalone program. Crashes typically require that the offending R process be killed using Task Manager before the program can be restarted...not ideal for end-users. In practice, this is exceedingly rare as long as basic best practices are followed.

* I normally copy finalized shiny code intended to run standalone over to a separate applications directory. Create a new folder in your applications working directory. In my case this would be `C:\data\RStudio\Applications`. For the `flight_proof` application used as an example below, the path would be: `C:\data\RStudio\Applications\flight_proof`  

* Download the latest version of [R-Portable](https://sourceforge.net/projects/rportable/files/R-Portable/) to the new folder. Then double-click the (currently) `R-Portable_3.5.2.paf.exe` to run. This will create the `R-Portable` folder in your new application folder. No admin privileges are needed to install `R-Portable`.  

* If a 32-bit version of R is needed to run your application, for example when a 32-bit MS Access database needs to be accessible, do not click `Yes` to the `Run R Portable` checkbox during the installation process...as the default assumes you want 64-bit R. 

* Install needed libraries: Normally you would install any needed packages by running the `Rgui.exe` located at `R-Portable\App\R-Portable\bin\x64`. For the less common cases where 32-bit R is needed I normally navigate to the `R-Portable\App\R-Portable\bin\i386` folder and then click on `Rgui.exe` to open the RGUI. I then install any libraries using 32-bit R. It may not make any difference either way.  

* Using the Rgui (whether 64-bit or 32-bit) open the `library_install_pak.R` script located in the [refreeze](https://github.com/arestrom/refreeze) repository. A copy is also normally kept in the `RStudio\Applications` folder. Edit as needed to add required libraries. This script uses the fabulous new [pak](https://cran.r-project.org/web/packages/pak/index.html) package to ensure all dependencies are installed along with the primary libraries in your `library()` calls. Prior to using the `pak` functions, getting all required libraries loaded could be a long and tedious process of testing, failing, and installing yet another undetected package dependency. 

* After opening the `library_install_pak.R` script in the `R-Portable` Rgui, highlight the relevant lines needed to install needed packages. Then click on `Run line or selection` icon. Enter `Yes` when asked to set up `pak` for initial use. You will be asked to say `Yes` on more time to install packages after they have been downloaded. You may need to scroll down in the R console occasionally to see progress as packages are installed.

* Copy your `ui.R`, `server.R`, and `global.R` shiny scripts, along with any other scripts needed for your application, into a folder named `shiny` in the top-level directory of your application. For example: `C:\data\RStudio\Applications\flight_proof\shiny`. There should also be a `www` subfolder containing any images, rmarkdown .rmd files, etc. 

* Make sure and edit the `server.R` file to uncomment the `stopApp()` function at the very bottom of the `server.R` script. This is needed to make sure the `R.exe` is properly closed when the application exits. The function looks like this:

```{r, eval = FALSE}
# close the R session when the browser closes
  session$onSessionEnded(function() {
    stopApp()
    q("no")
  })
```

* The [refreeze](https://github.com/arestrom/refreeze) repository contains example materials needed for the next steps. Copy the `run.vbs`, `runShinyApp.R`, and `FlightProof.ico` from the repository into the top level of your app folder. You should also copy the `flight_proof.iss` script that will be needed to create the setup executable. I normally keep a copy of all application `.iss` scripts in the application parent directory, in my case `C:\data\RStudio\Applications`.  

* Rename the `FlightProof.ico` as needed, or find another icon to use. There should be no need to edit the other files. You will only need to edit the `run.vbs` if 32-bit R is needed. In that case the text of the `run.vbs` script will be `R-Portable\App\R-Portable\bin\i386\R.exe` instead of `R-Portable\App\R-Portable\bin\x64\R.exe`. 

### Create Windows installer

* Download and install [Inno Setup](http://www.jrsoftware.org/isinfo.php). Then edit a copy of the `flight_proof.iss` Inno Setup script, and rename as needed to your new application name, for example, `app_name.iss`. 

* Double-click on the `app_name.iss` script to open `Inno Setup`. In the `AppID` field, highlight all but the first curly brace, and under `Tools` click on the `Generate GUID` option. This will overwrite the section with a new GUID. Edit the other fields as needed to specify things such as the application name and version number, the source directory for the application files, and the destination directory for the setup executable. Make sure that the `PrivilegesRequired` field is set to `none`. This will ensure that end-users can run the setup executable without needing administrative privileges. In practice I have found that the setup dialogue may still ask for an administrative login in order to run the installer, but your normal login account name and password will work. 

* To create the application setup installer, click on `Build – Compile`. This will bundle up all materials needed for you application and create an executable installer in your application directory. Your application, along with the current version of R and all needed packages, are now frozen in time. 

### In case of failure  

* If after installing the application, it crashes, or fails to load, navigate to the `out.txt` folder in the App directory that the installer created, for example: `C:\data\Intertidal\Apps\FlightProof\out.txt`. The `out.txt` file will log the first fatal error encountered. In the case of missing dependencies it will identify any packages that are still missing. Each missing dependency will trigger a crash, so in the past, before using the `pak` functions for library installation, multiple iterations were typically needed to identify all missing packages.  

* In case the application crashes or fails to load, you will also most likely need to kill any R processes still active using task manager. You will then need to fix the source of the crash, usually a missing package dependency, and recompile the application. 


